---
title: "Combined analysis of multiple models"
format:
    html:
        embed-resources: true
        default-image-extension: svg
project:
    execute_dir: project
editor: visual
---

## Load data

```{r read_data}
#| include: false

library(tidyverse)
library(targets)
library(phyloseq)
library(microViz)
library(brms)
library(posterior)
library(ggdist)
library(ggVennDiagram)
library(speedyseq)
library(patchwork)
library(metacoder)
tar_source()

ps_its <- tar_read(phyloseq_its, store = here::here("_targets")) |> 
  ps_mutate(marker = "ITS")
tax_table(ps_its) <- tax_table(ps_its)[,1:7]
ps_16s <- tar_read(phyloseq_16s, store = here::here("_targets")) |> 
  ps_mutate(marker = "16S") |> 
  fix_ps_tax("data/asvs.taxonomy.sativa.tsv")

tar_load(site_df)
```

## Taxonomy - metacoder heat trees

```{r}
#| echo: false
#| warnings: false

its_mc <- parse_phyloseq(ps_its)
its_mc$data$tax_abund <- calc_taxon_abund(its_mc, "otu_table")

rrna_mc <- parse_phyloseq(ps_16s)
rrna_mc$data$tax_abund <- calc_taxon_abund(rrna_mc, "otu_table")

its_mc$data$tax_occ <- calc_n_samples(its_mc, 
                                      "tax_abund", 
                                      groups = sample_data(ps_its)$Tissue,
                                      cols = rownames(sample_data(ps_its))
                                      )
rrna_mc$data$tax_occ <- calc_n_samples(rrna_mc, 
                                       "tax_abund", 
                                       groups = sample_data(ps_16s)$Tissue,
                                      cols = rownames(sample_data(ps_16s))
                                      )
(its_tree <- its_mc |> 
  filter_taxa(taxon_ranks == "Class", supertaxa = TRUE) |> 
  heat_tree(node_label = taxon_names,
            node_label_size_range = c(0.02, 0.05),
            node_size = n_obs,
            node_color = n_obs,
            title = NULL,
            initial_layout = "reingold-tilford", layout = "reingold-tilford",
            # node_size_axis_label = "Number of ASVs",
            node_color_axis_label = "Number of ASVs",
            make_edge_legend = FALSE
            ) +
    theme(rect = element_rect(fill = "transparent")))

rrna_tree <- rrna_mc |> 
  filter_taxa(taxon_ranks == "Class", supertaxa = TRUE) |> 
  heat_tree(node_label = taxon_names,
            node_label_size_range = c(0.02, 0.05),
            node_size = n_obs,
            node_color = n_obs,
            title = NULL,
            initial_layout = "reingold-tilford", layout = "reingold-tilford",
            # node_size_axis_label = "Number of OTUs",
            node_color_axis_label = "Number of ASVs"
            ) +
    theme(rect = element_rect(fill = "transparent"))

tax_trees <- rrna_tree + its_tree + plot_annotation(tag_levels = "a")
ggsave("../Papers/Figures/tax_trees.png", width = 180, height = 90, units = "mm")
```

## Community composition metrics

How does the community composition differ between the different tissue types? Comparing both ordination and richness of ASVS:

```{r}
ps <- merge_phyloseq(ps_its, ps_16s) |> 
  ps_mutate(SampleID = paste(Tissue, Site, Tree, sep = "-")) |> 
  speedyseq::merge_samples2(group = "SampleID")

asvs_leaf <- ps |> ps_filter(Tissue == "Leaf") |> taxa_names()
asvs_stem <- ps |> ps_filter(Tissue == "Stem") |> taxa_names()
asvs_soil <- ps |> ps_filter(Tissue == "Soil") |> taxa_names()

venn_df <- process_data(
  Venn(list(Leaf = asvs_leaf,
            Stem = asvs_stem,
            Soil = asvs_soil))
)

venn <- ggplot() +
  geom_polygon(aes(X, Y, fill = id, group = id), data = venn_regionedge(venn_df), show.legend = FALSE) +
  geom_path(aes(X, Y, group = id), colour = "black", data = venn_setedge(venn_df), show.legend = FALSE) +
  geom_text(aes(X, Y, label = name), data = venn_setlabel(venn_df), size = 2) +
  geom_label(aes(X, Y, label = count), fontface = "bold", size = 2, data = venn_regionlabel(venn_df)) +
  scale_fill_manual(values = c("1" = "olivedrab3", 
                               "2" = "bisque2", 
                               "3" = "salmon4", 
                               "1/2" = "#C4D175",
                               "1/3" = "#938D36",
                               "2/3" = "#BD9178",
                               "1/2/3" = "#b1a561")) +
  coord_equal() +
  # scale_x_continuous(expand = expansion(mult = .2)) +
  theme_void() + 
  theme(legend.position = "none", 
        panel.background = element_rect(colour = "black", fill = "white", linewidth = 1))

venn
```

Leaf is by far the least diverse tissue, with the fewest total ASVs. The overlap in ASVs appears to depend on the proximity of the tissue types to one another - leaf and stem, and stem and soil, share more ASVs than leaf and soil. There are very few global ASVs.

Looking at the ordination shows a very similar pattern:

```{r}
ord <- ordinate(ps, method = "NMDS", distance = "horn", trymax = 100, parallel = 8)
ord_plot <- plot_ordination(ps, ord, justDF = TRUE) |> 
  ggplot(aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(colour = Tissue), size = 1) +
  scale_colour_manual(values = c(Leaf = "olivedrab3", Stem = "bisque2", Soil = "salmon4")) +
  lims(x = c(-4, 4)) +
  # labs(x = NULL, y = NULL) +
  theme_bw() + 
  theme(legend.position = "none")

ord_inset <- ord_plot + inset_element(venn, left = 0.6, bottom = 0.6, top = 0.99, right = 0.99)

ord_plot
```

The ordination seems to suggest that leaf samples are more tightly clustered together - we can quantify the average similarity among samples from each tissue type to check this. Here, using the horn index, where higher = more similarity among samples:

```{r}
leaf_sim <- rvar(as.vector(1 - vegan::vegdist(t(otu_table(ps_filter(ps, Tissue == "Leaf"))), distance = "horn")))
stem_sim <- rvar(as.vector(1 - vegan::vegdist(t(otu_table(ps_filter(ps, Tissue == "Stem"))), distance = "horn")))
soil_sim <- rvar(as.vector(1 - vegan::vegdist(t(otu_table(ps_filter(ps, Tissue == "Soil"))), distance = "horn")))

sim_df <- tibble(Tissue = c("Leaf", "Stem", "Soil"), 
                 horn = c(leaf_sim, stem_sim, soil_sim),
                 num = 1:3
                 )

ggplot(sim_df, aes(x = fct_reorder(Tissue, num), ydist = horn)) + 
  stat_dots(aes(colour = Tissue), layout = "weave") +
  scale_colour_discrete(type = c("olivedrab3", "salmon4", "bisque2")) +
  # scale_colour_manual(values = c(Leaf = "olivedrab3", Stem = "bisque2", Soil = "salmon4")) +
  labs(x = "Tissue", y = "Pairwise similarity") + 
  theme_bw() + 
  theme(legend.position = "bottom")
```

## Association of microbial class with tissue

We see that each tissue has drastically different microbial communities - but which taxonomic groups are driving this?

```{r}
ps_its_class <- tax_glom(ps_its, taxrank = rank_names(ps_its)[3])
ps_16s_class <- tax_glom(ps_16s, taxrank = rank_names(ps_16s)[3])
taxa_names(ps_its_class) <- ps_its_class@tax_table[,3]
taxa_names(ps_16s_class) <- ps_16s_class@tax_table[,3]

ps_its_class_long <- psmelt(ps_its_class) |> 
  mutate(TotalReads = sum(Abundance), .by = "Sample") |> 
  mutate(marker = "ITS")

ps_16s_class_long <- psmelt(ps_16s_class) |> 
  mutate(TotalReads = sum(Abundance), .by = "Sample") |> 
  mutate(marker = "16S")

contrasts <- function(rvar_vec) {
  combinations <- combn(1:length(rvar_vec), m = 2)
  
  contrasts <- map(1:length(rvar_vec), \(x)
      rvar_vec[combinations[1,x]] - rvar_vec[combinations[2,x]]
      ) |> list_c()
}

mod <- cmdstanr::cmdstan_model("Stan/bb_mod.stan")

make_bb_mod_data <- function(df, category) {
  cat <- factor(pull(df, category)) |> 
    fct_relevel(c("Leaf", "Stem", "Soil"))

  list(
    N = nrow(df),
    N_cat = length(unique(cat)),
    Y = df$Abundance,
    Total = df$TotalReads,
    Categories = cat,
    cor = matrix(c(1, 0.3, 0.3, 1), ncol = 2),
    sigma = c(1, 1)
  )
}

ps_full <- bind_rows(ps_its_class_long, ps_16s_class_long) |> 
  filter(TotalReads != 0) |>
  mutate(Tissue_num = as.numeric(fct_relevel(Tissue, c("Leaf", "Stem", "Soil")))) |> 
  nest(.by = c(Kingdom, Phylum, Class, marker)) |> 
  mutate(standata = map(data, \(x) make_bb_mod_data(x, "Tissue")),
         model = map(standata, 
                     \(x) mod$sample(data = x, chains = 4, parallel_chains = 4))
         )

ps_full2 <- ps_full |> 
  mutate(
    divergences = list_c(map(model, \(x) sum(x$diagnostic_summary()$num_divergent))),
    converged = list_c(map(model, \(x) as_draws(x) |> summarise_draws() |> pull(rhat) |> {\(y) !any(y > 1.05)}())),
    epred = map(model, \(x) as_draws_rvars(x) |> pluck("epred")),
    pred = map(model, \(x) as_draws_rvars(x) |> pluck("ppred")),
    contrasts_epred = map(epred, \(x) contrasts(x)),
    contrasts_pred = map(pred, \(x) contrasts(x)),
    any_sig_contrasts = map(contrasts_epred, \(x) any(mean(x >= -0.05 & x <= 0.05) < 0.95)) |> list_c()
  ) |>
  pivot_longer(cols = c(epred, contrasts_epred, pred, contrasts_pred), 
               names_to = "PP", values_to = "pred") |>
  unnest_longer(pred) |>
  mutate(Contrast = c("Leaf - Stem", "Leaf - Soil", "Stem - Soil"),
         Tissue = c("Leaf", "Stem", "Soil"),
         .by = c(Class, marker, PP)
  )

n_sig <- ps_full2 |> 
  filter(PP == "epred", any_sig_contrasts == TRUE) |> 
  pull(Class) |> unique() |> length()

(props <- ps_full2 |> 
  filter(PP == "epred", any_sig_contrasts == TRUE) |> 
  mutate(label = glue::glue("{Kingdom}: {Class}")) |> 
  ggplot(aes(xdist = pred, y = fct_rev(label), colour = Tissue)) + 
  stat_pointinterval(position = position_dodge(width = 0.8), size = 0.1) + 
  geom_hline(yintercept=seq(1.5, n_sig - 0.5, 1),
             linetype = "dashed", colour = "grey60") +
  scale_colour_manual(values = c(Leaf = "olivedrab3", Stem = "bisque2", Soil = "salmon4")) +
  # facet_grid(rows = vars(label), scales = "free_y") +
  labs(x = "Mean relative abundance", y = NULL) + 
  theme_bw() + 
  theme(legend.position = "none",
        axis.text = element_text(size = 5),
        strip.background = element_blank(),
        strip.text = element_blank()
        # panel.spacing.y = unit(0, "lines")
        ))
```

## Kingdom analysis

```{r}
deotherise <- function(df){
  other_count <- df |> filter(Kingdom == "Other") |> pull(Count)
  df |> mutate(Total = Total - other_count) |> 
    filter(Kingdom != "Other")
}

mg_md <- tar_read(tree_df) |> 
  select(Site, Tree, Severity, Leaf = MG_leaf, Stem = MG_stem, Soil = MG_soil) |> 
  pivot_longer(cols = c(Leaf, Stem, Soil), names_to = "Tissue", values_to = "SampleID")
mg_kingdom <- read_csv("data/mg_kingdom_summary.csv") |> 
  left_join(mg_md) |> 
  mutate(Tissue_num = as.numeric(fct_relevel(Tissue, c("Leaf", "Stem", "Soil"))),
         Disease = ifelse(Severity == "Control", 0, 1)) |> 
  nest(.by = SampleID) |> 
  rowwise() |> 
  mutate(data = list(deotherise(data))) |> 
  unnest(cols = c(data))
  
mg_kingdom_tissue_mods <- mg_kingdom |> 
  nest(.by = c(Kingdom)) |> 
  mutate(standata = map(data, \(x) make_bb_mod_data(x, "Tissue_num")),
         model = map(standata, 
                     \(x) mod$sample(data = x, chains = 4, parallel_chains = 4)),
         divergences = list_c(map(model, \(x) sum(x$diagnostic_summary()$num_divergent))),
         converged = list_c(map(model, \(x) as_draws(x) |> summarise_draws() |> pull(rhat) |> {\(y) !any(y > 1.05)}())),
         epred = map(model, \(x) as_draws_rvars(x) |> pluck("epred")),
         pred = map(model, \(x) as_draws_rvars(x) |> pluck("ppred"))
  ) |>
  pivot_longer(cols = c(epred, pred), 
               names_to = "PP", values_to = "pred") |>
  unnest_longer(pred) |>
  mutate(Tissue = c("Leaf", "Stem", "Soil"),
         .by = c(Kingdom, PP)
  )

(kingdom_props <- mg_kingdom_tissue_mods |> 
  filter(PP == "pred", Kingdom != "Other") |>
  mutate(label = Kingdom) |> 
  ggplot(aes(xdist = pred / 10000, y = Tissue, colour = Kingdom)) + 
  stat_pointinterval(position = position_dodge(width = 0.8), size = 4) + 
  geom_hline(yintercept=seq(1.5, n_sig - 0.5, 1),
             linetype = "dashed", colour = "grey60") +
  # scale_colour_manual(values = c(Leaf = "olivedrab3", Stem = "bisque2", Soil = "salmon4")) +
  # facet_grid(rows = vars(label), scales = "free_y") +
  labs(x = "Predicted relative abundance", y = NULL) + 
  theme_bw())

pred_dat <- expand_grid(
  Total = 10000,
  groundfrost = 0,
  hurs = 0,
  pv = 0,
  rainfall = 0,
  sfcWind = 0,
  sun = 0,
  tas = 0,
  Ntot = 0,
  SOx = 0,
  CaMg = 0,
  Disease = c(0, 1)
)

mg_kingdom_disease <- mg_kingdom |>
  left_join(tar_read(site_df)) |> 
  mutate(across(groundfrost:Cl, \(x) scale(x)[,1])) |> 
  nest(.by = c(Kingdom, Tissue)) |> 
  mutate(model = map(data, \(x) brm(
    bf(Count | trials(Total) ~ 0 + Intercept + (groundfrost + hurs + pv + rainfall + sfcWind + sun + tas + Ntot + SOx + CaMg + Disease),
       phi ~ 0 + Intercept,
       family = "beta_binomial"),
    prior = c(
      prior(normal(0, 1), class = "b")
      ),
    backend = "cmdstanr",
    chains = 4,
    cores = 4,
    data = x
    )
  ),
  pred = map(model, \(x) pred_dat |> 
               tidybayes::add_epred_rvars(x) |> 
               tidybayes::add_predicted_rvars(x)
             )
  ) |> 
  unnest(pred) |> 
  select(Kingdom, Tissue, Disease, pred = .epred) |> 
  mutate(pred = pred / 10000 * 100) |> 
  pivot_wider(names_from = Disease, values_from = pred) |> 
  mutate(diff = `1` - `0`)

ggplot(mg_kingdom_disease, aes(xdist = diff, 
                               y = Kingdom)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey80") +
  stat_pointinterval() +
  facet_grid(rows = vars(Tissue), scales = "free") +
  labs(x = "Change in relative abundance in diseased trees (%)") +
  theme_bw()
```

## Alpha diversity

We are also interested in alpha diversity metrics, and how these might vary as a function of variables of interest - in particular, AOD.

First, let's look at the raw data. First, species richness:

```{r}
raw_alpha_div <- map(c("ITS", "16S"), \(x)
                     map(c("Leaf", "Stem", "Soil"), \(y)
                         get_per_sample_raw_richness(x, y, 
                           measures = c("Observed", "Chao1", "ACE", "Shannon", "Simpson", "InvSimpson")))
                     ) |> list_flatten() |> list_rbind()

ggplot(raw_alpha_div, aes(x = Tissue, y = Observed)) + 
  geom_dots(aes(colour = Disease, fill = Disease), layout = "swarm") + 
  scale_colour_hue(direction = -1) +
  scale_fill_hue(direction = -1) +
  facet_grid(rows = vars(Marker)) +
  theme_linedraw()
```

There might be a very slight decrease in richness in diseased trees for some markers/compartments? Let's run a quick comparison:

```{r}
richness_mod_nb_shape <- brm(
  formula = bf(
    Observed ~ 0 + Intercept + Tissue*Marker*Disease,
    shape ~ 0 + Intercept + Tissue * Marker
    ),
  family = "negbinomial",
  prior = c(prior(normal(0, 1), class = "b")),
  data = raw_alpha_div,
  backend = "cmdstan",
  cores = 4,
  chains = 4
)

newdata <- expand.grid(Tissue = c("Leaf", "Stem", "Soil"),
                       Marker = c("ITS", "16S"),
                       Disease = c("Control", "DiseaseAOD"),
                       TotalReads = 10000)

richness_mod_df <- newdata |> 
  tidybayes::add_epred_draws(richness_mod_nb_shape) |> 
  group_by(Tissue, Marker, Disease) |> 
  summarise(pred = rvar(.epred)) |> 
  pivot_wider(names_from = Disease, values_from = pred) |>
  mutate(Diff = DiseaseAOD - Control) |>
  mutate(Model = "ObsRichness")

richness_mod_df |>
  ggplot(aes(x = Tissue)) +
  stat_pointinterval(aes(ydist = Diff, colour = Marker), position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_grid(cols = vars(Tissue), scales = "free") +
  labs(y = "Species richness") +
  theme_linedraw()
```

Using the best fitting model (`richness_mod_nb_shape`), there appears to be a decrease in observed richness in diseased trees to about 85 to 90% of healthy richness in leaf and stem, and an increase in richness of about 120% in soil fungi.

Let's compare this to the patterns we see from the occupancy model. We can do this in two ways, 1: we can average across all actual samples, and get the specific differences in richness observed between the models, 2: assume a hypothetical site with all species present, and estimate the difference assuming a diseased and healthy tree at that site.

```{r}
richnesses_occ_avg <- map(c("ITS", "16S"), \(x)
              map(c("Leaf", "Stem", "Soil"), \(y)
                  calculate_average_richness_differences(x, y))
              ) |> list_flatten() |> list_rbind() |> 
  mutate(Model = "Occ_Avg")

richnesses_occ_hyp <- map(c("ITS", "16S"), \(x)
              map(c("Leaf", "Stem", "Soil"), \(y)
                  model_richness_disease_differences(x, y, n_env_vars = 10, chunk_size = 20))
              ) |> list_flatten() |> list_rbind() |> 
  mutate(Disease = ifelse(Disease == 0, "Control", "DiseaseAOD")) |> 
  pivot_wider(names_from = "Disease", values_from = "pred") |> 
  mutate(Diff = DiseaseAOD - Control) |> 
  mutate(Model = "Occ_Hyp")

richness_comp_plot <- bind_rows(richness_mod_df, 
                                richnesses_occ_hyp) |> 
  filter(Model == "Occ_Hyp") |> 
  ggplot(aes(y = Tissue, xdist = Diff, colour = Marker)) + 
  stat_pointinterval(position = position_dodge(width = 0.3), 
                     size = 3, linewidth = 1, .width = 0.95) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  # facet_grid(rows = vars(Marker)) +
  scale_colour_discrete(type = c(`16S` = "slateblue3", ITS = "tomato1")) +
  labs(colour = "Marker", x = "Difference in species richness (AOD - Control)", y = NULL) +
  theme_linedraw() +
  theme(legend.position = "bottom")

richness_comp_plot2 <- richnesses_occ_avg |> 
  ggplot(aes(y = Tissue, xdist = Diff, colour = Model)) + 
  stat_pointinterval(position = position_dodge(width = 0.3)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_grid(rows = vars(Marker)) +
  scale_colour_hue(direction = -1) + 
  labs(colour = "Marker gene", x = "Difference in species richness (AOD - Control)", y = NULL) +
  theme_linedraw() +
  theme(legend.position = "bottom")
```

Matches the expected pattern from the observed data - though the CIs are wider. Nice!

Finally, we want to know about how much non-detection affects the estimated species richness:

```{r}
occ_richness_site_df <- map(c("ITS", "16S"), \(x)
                            map(c("Leaf", "Stem", "Soil"), \(y)
                                make_richness_df(x, y, level = 1))
                            ) |> list_flatten() |> list_rbind()

ntaxas <- function(marker, tissue) {
  ps <- tar_read_raw(glue::glue("microbiome_tj_corrected_{marker}_{tissue}"))
  tibble(Marker = marker, 
         Tissue = tissue,
         NTaxTotal = ntaxa(ps))
}

ntaxas_df <- map(c("ITS", "16S"), \(x)
                            map(c("Leaf", "Stem", "Soil"), \(y)
                                ntaxas(x, y))
                            ) |> list_flatten() |> list_rbind()

occ_richness_site <- occ_richness_site_df |> 
  mutate(Ntrees = as.factor(ifelse(AODProx == "AOD", 20, 5))) |> 
  left_join(ntaxas_df) |> 
  mutate(n_abs = NTaxTotal - Richness_mod) |> 
  summarise(Richness_obs_diff = rvar_mean(Richness_obs_diff, na.rm = TRUE),
            Richness_abs = rvar_mean(n_abs, na.rm = TRUE),
            .by = c(Tissue, Marker)) 

occ_richness_site_ntrees <- occ_richness_site_df |> 
  mutate(Ntrees = as.factor(ifelse(AODProx == "AOD", 20, 5))) |> 
  left_join(ntaxas_df) |> 
  mutate(n_abs = NTaxTotal - Richness_mod) |> 
  summarise(Richness_obs_diff = rvar_mean(Richness_obs_diff, na.rm = TRUE),
            Richness_abs = rvar_mean(n_abs, na.rm = TRUE),
            .by = c(Tissue, Marker, Ntrees)) 

ggplot(occ_richness_site, aes(x = Marker, ydist = Richness_obs_diff)) +
  stat_pointinterval(na.rm = TRUE) +
  scale_colour_discrete() +
  facet_grid(cols = vars(Tissue)) +
  theme_light()

ggplot(occ_richness_site_ntrees, aes(x = Marker, ydist = Richness_obs_diff, colour = Ntrees)) +
  stat_pointinterval(na.rm = TRUE) +
  scale_colour_discrete() +
  facet_grid(cols = vars(Tissue)) +
  theme_light()
```

## Plot full SDs for all models

```{r}
env_vars <- c("Intercept", "Groundfrost", "Relative humidity", "Vapour pressure", "Rainfall", "Windspeed", "Sunshine", "Air temperature", "N deposition", "Sulphur deposition", "Ion deposition")
sample_vars <-  "Disease"

sigmas_df <- map(c("ITS", "16S"), \(x)
                 map(c("Leaf", "Stem", "Soil"), \(y)
                     map(0:3, \(z) 
                         make_sigmas_df(x, y, level = z, env_vars = env_vars, sample_vars = sample_vars))
                     |> list_rbind()
                 ) |> list_rbind()
) |> list_rbind() |> 
  mutate(org = case_when(
    marker == "ITS" ~ "Fungi",
    marker == "16S" ~ "Bacteria",
    .default = NA
  ),
  level_name = case_when(
    level == 0 ~ "Total",
    level == 1 ~ "Site",
    level == 2 ~ "Sample",
    level == 3 ~ "Reads",
    .default = NA
  )
  )

sigmas_plot <- sigmas_df |>
  filter(level != 0, var != "Intercept") |> 
  # filter(!(level == 1 & var == "Disease")) |> 
  ggplot(aes(y = fct_rev(var))) +
  ggdist::stat_ccdfinterval(aes(xdist = sigma, 
                                fill = as.factor(level),
                                slab_alpha = after_stat(f)), 
                            fill_type = "gradient",
                            position = position_dodge(width = 0.9),
                            thickness = 1
                            ) +
  # scale_fill_manual(values = c("slateblue3", "tomato1")) +
  facet_grid(cols = vars(fct_relevel(tissue, c("Leaf", "Stem", "Soil"))),
             rows = vars(org)) +
  labs(y = NULL, x = "CCDF of variance parameter", fill = NULL) +
  guides(slab_alpha = "none") +
  theme_bw() +
  theme(legend.position = "bottom")

sigmas_bacteria_plot <- sigmas_df |>
  filter(org == "Bacteria", level_name != "Total", var != "Intercept") |> 
  filter(!(level == 1 & var == "Disease")) |> 
  ggplot(aes(y = fct_rev(var))) +
  ggdist::stat_ccdfinterval(aes(xdist = sigma^2, 
                                fill = level_name,
                                slab_alpha = after_stat(f)), 
                            fill_type = "gradient",
                            position = position_dodge(width = 0.9),
                            thickness = 1
                            ) +
  facet_grid(cols = vars(fct_relevel(tissue, c("Leaf", "Stem", "Soil")))) +
  labs(y = NULL, x = "CCDF of variance parameter", fill = NULL) +
  guides(slab_alpha = "none") +
  theme_bw() +
  theme(legend.position = "bottom")

sigmas_fungi_plot <- sigmas_df |>
  filter(org == "Fungi", level_name != "Total", var != "Intercept") |> 
  filter(!(level == 1 & var == "Disease")) |> 
  ggplot(aes(y = fct_rev(var))) +
  ggdist::stat_ccdfinterval(aes(xdist = sigma^2, 
                                fill = level_name,
                                slab_alpha = after_stat(f)), 
                            fill_type = "gradient",
                            position = position_dodge(width = 0.9),
                            thickness = 1
                            ) +
  facet_grid(cols = vars(fct_relevel(tissue, c("Leaf", "Stem", "Soil")))) +
  labs(y = NULL, x = "CCDF of variance parameter", fill = NULL) +
  guides(slab_alpha = "none") +
  theme_bw() +
  theme(legend.position = "bottom")

sigmas_fungi_plot / sigmas_bacteria_plot + 
  plot_layout(axes = "collect", guides = "collect", axis_titles = "collect") & 
  theme(legend.position = "bottom")
#ggsave("sigmas_occ_plot.svg", sigmas_plot, width = 8, height = 5)
```

## Environmental slopes

```{r}
env_vars <- tar_read(env_vars)
env_var_names <- c("Groundfrost", "Relative humidity", "Vapour pressure", "Rainfall", "Windspeed", "Sunshine", "Air temperature", "N deposition", "Sulphur deposition", "Ion deposition")
quantile_slopes_df <- map(c("ITS", "16S"), \(x)
                 map(c("Leaf", "Stem", "Soil"), \(y)
                     species_slopes_quantiles(x, y, level = 3, env_vars = env_vars, var_names = env_var_names, site_data = site_df)
                     )
                 ) |> list_flatten() |> list_rbind()

quantile_slopes_plot <- quantile_slopes_df |> 
  filter(level == 1) |> 
  pivot_wider(names_from = ci, values_from = slope) |> 
  mutate(x = (seq * sd) + mean,
         lci_y = exp(lci * seq),
         uci_y = exp(uci * seq)) |> 
  ggplot(aes(x = x, ymin = lci_y, ymax = uci_y, fill = Marker, colour = Marker)) +
  geom_ribbon(alpha = 0.5) +
  scale_fill_manual(values = c("slateblue3", "tomato1")) +
  scale_colour_manual(values = c("slateblue3", "tomato1")) +
  facet_grid(cols = vars(var_names), 
             rows = vars(fct_relevel(Tissue, c("Leaf", "Stem", "Soil"))), 
             scale = "free_x") + 
  labs(x = NULL, y = "Fold change in site-level occupancy probability") +
  coord_cartesian(ylim = c(0, 20)) +
  theme_linedraw() +
  theme(legend.position = "bottom",
        text = element_text(size = 7),
        title = element_text(size = 7),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Disease-associated microbes

We want to identify a list of taxa that are putatively associated with plant health status. As we are using a Bayesian approach, we ideally want to use a ROPE test to pull these out, as there's not a clear expectancy that any effect is exactly 0. Instead, we need to define a range of values that we consider "practically zero", and see what proportion of each effect is in that range. For log2 fold changes, I think a range of \[-0.18, 0.18\] is quite a reasonable 'null' as this equates to a range of fold changes from 0.88x to 1.13x. We then need to chose a confidence threshold - if over X% is in the ROPE, we say the effect is not 'significant'. Here, I am using an 11% confidence threshold.

```{r}
disease_effects <- map(c("ITS", "16S"), \(x)
              map(c("Leaf", "Stem", "Soil"), \(y)
                  disease_intervals(x, y, chunk_size = 20))
              ) |> list_flatten() |> list_rbind() |>
    rowwise() |> 
    mutate(ASV_numeric = as.numeric(ASV_numeric),
           diff_abs = DiseaseAOD - Control,
           lci_abs = quantile(diff_abs, probs = 0.025),
           med_abs = quantile(diff_abs, probs = 0.5),
           uci_abs = quantile(diff_abs, probs = 0.975)) |> 
    ungroup()

disease_effects <- disease_effects |> 
  mutate(diff_lfc = log2(DiseaseAOD) - log2(Control),
    rope = bayestestR::rope(diff_lfc, ci = 1, range = c(-0.1, 0.1))$ROPE_Percentage * 100)

microbes <- filter(disease_effects, rope <= 2.5)
disease_microbes <- filter(microbes, med_abs > 0)
health_microbes <- filter(microbes, med_abs < 0)

(tax_plot <- plot_disease_intervals(microbes) + theme(legend.position = "bottom"))

taxonomy <- select(disease_effects, ASV_ID:Species_exact)

reps_health <- representation_likelihood(tax_level = "Order",
                                  sig_list = health_microbes,
                                  taxa_list = taxonomy,
                                  stan_file = "Stan/hyper_intercept.stan") |>
  mutate(Assoc = "Asymptomatic")

reps_disease <- representation_likelihood(tax_level = "Order",
                                  sig_list = disease_microbes,
                                  taxa_list = taxonomy,
                                  stan_file = "Stan/hyper_intercept.stan") |>
  mutate(Assoc = "AOD")

reps <- bind_rows(reps_health, reps_disease) |>
  filter(n_bag != n_pick) |> 
  filter(n_pick != 0) |> 
  mutate(median = median(alpha),
         rope = bayestestR::rope(alpha, ci = 1, range = c(-0.18, 0.18))$ROPE_Percentage)

tax_kingdoms <- select(taxonomy, Kingdom, Order) |> distinct()

res <- filter(reps, rope < 0.025) |>
  filter(Order != "") |>
  left_join(tax_kingdoms)

(tax_assoc_plot <- res |> 
  mutate(tax = glue::glue("{Kingdom}: {Order}")) |> 
  ggplot(aes(xdist = alpha, y = fct_reorder(tax, median), colour = Assoc)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey70") +
  stat_pointinterval(.width = c(.0, .89)) +
  scale_colour_discrete(type = c(AOD = "indianred2", Asymptomatic = "steelblue2")) +
  labs(y = "Class", x = "log odds pulling clade from bag", colour = "Disease") +
  theme_linedraw() +
  theme(legend.position = "bottom"))
```

------------------------------------------------------------------------

## Tax figure

```{r}
func_conts <- qs::qread("functions/tissue_contrasts.qs") |> 
  ggplot(aes(y = forcats::fct_rev(lbl),#forcats::fct_reorder(lbl, log2FoldChange, .fun = mean), 
           x = log2FoldChange, 
           xmin = log2FoldChange - ci, 
           xmax = log2FoldChange + ci,
           colour = coef)) +
  geom_pointrange(size = 0.1) +
  geom_vline(xintercept = 0, colour = "grey70", linetype = "dashed") +
  facet_grid(cols = vars(coef)) +
  labs(x = "log2 fold change", y = "", colour = "") +
  theme_linedraw() +
  theme(legend.position = "bottom",
        axis.text = element_text(size = 5),
        axis.title = element_text(size = 7))

func_ord <- qs::qread("functions/tissue_ord.qs") |> 
  ggplot(aes(x = NMDS1, y = NMDS2, colour = Tissue)) +
  geom_point(size = 1) +
  scale_colour_discrete(type = c(Leaf = "olivedrab3", Stem = "bisque2", Soil = "salmon4")) +
  theme_linedraw() +
  theme(legend.position = "bottom")
# 
# wrap_plots(free(rrna_tree), free(its_tree), free(props), ord_inset, func_conts, func_ord, ncol = 2) +
# plot_annotation(tag_levels = "a") &
#   theme(text = element_text(size = 6),
#         title = element_text(size = 7))

wrap_plots(free(props), ord_inset, func_conts, func_ord, ncol = 2) +
  plot_layout(widths = c(0.4, 0.6)) +
  plot_annotation(tag_levels = "a") &
  theme(text = element_text(size = 6),
        title = element_text(size = 7))
ggsave("../Papers/Figures/tissue_comparisons.png", dpi = 300, width = 180, height = 180, units = "mm")
```

## env figure

```{r}
func_env_slopes <- qs::qread("functions/env_slopes.qs") |> 
  mutate(coef_full = case_when(
      coef_full == "CaMg" ~ "Ion deposition",
      coef_full == "Nitrogen deposition" ~ "N deposition",
      coef_full == "Bright sunshine" ~ "Sunshine",
      .default = coef_full
    ),
    prac_sig = ifelse(abs(log2FoldChange) >= 0.1, "1", "0")
  ) |> 
  ggplot(aes(x = x, y = 2^pred)) + 
  geom_line(aes(group = comb, colour = prac_sig), alpha = 0.8) + 
  scale_colour_discrete(type = c("grey70", "black")) +
  facet_grid(cols = vars(coef_full), 
             rows = vars(fct_relevel(Tissue, c("Leaf", "Stem", "Soil"))), 
             scales = "free_x") +
  labs(x = NULL, y = "Fold change in GO term abundance") +
  coord_cartesian(ylim = c(0, 10)) +
  theme_linedraw() + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

((quantile_slopes_plot + labs(title = "Taxonomy")) / (func_env_slopes  + labs(title = "Function"))) +
  plot_annotation(tag_levels = "a") & 
  theme(text = element_text(size = 6),
        title = element_text(size = 7))

ggsave("../Papers/Figures/env_comparisons.png", dpi = 300, width = 180, height = 150, units = "mm")
```

## maps

```{r}
its_leaf_asvs <- tar_read(landscape_predictions_theta_ITS_Leaf)
rrna_stem_asvs <- tar_read(landscape_predictions_theta_16S_Stem)
rrna_stem_asvs <- tar_read(landscape_predictions_theta_16S_Soil)

map1 <- plot_raster(its_leaf_asvs, tar_read(landscape_df), tar_read(site_df), tar_read(microbiome_tj_corrected_ITS_Leaf), asv_tbl = tar_read(asv_table_ITS_Leaf), 34)
map2 <- plot_raster(rrna_stem_asvs, tar_read(landscape_df), tar_read(site_df), tar_read(microbiome_tj_corrected_16S_Stem), asv_tbl = tar_read(asv_table_16S_Stem), 726)
map3 <- plot_raster(rrna_stem_asvs, tar_read(landscape_df), tar_read(site_df), tar_read(microbiome_tj_corrected_16S_Soil), asv_tbl = tar_read(asv_table_16S_Soil), 34)
```

## disease figure

```{r}
disease_conts <- qs::qread("functions/disease_contasts.qs") |> 
  ggplot(aes(y = forcats::fct_reorder(lbl, log2FoldChange, .fun = mean), 
           x = log2FoldChange, 
           xmin = (log2FoldChange - ci), 
           xmax = (log2FoldChange + ci))) +
  geom_pointrange(size = 0.01, linewidth = 0.5) +
  geom_vline(xintercept = 0, colour = "grey70", linetype = "dashed") +
  labs(x = "Fold Change", y = "", colour = "") +
  theme_linedraw() +
  theme(legend.position = "bottom")

disease_ord <- qs::qread("functions/disease_ord.qs") |> 
  ggplot(aes(x = NMDS1, y = NMDS2, fill = Disease)) +
  geom_point(size = 3, shape = 21) +
  scale_fill_discrete(type = c(AOD = "indianred2", Asymptomatic = "steelblue2")) +
  theme_linedraw() +
  theme(legend.position = "none")

tax_assoc_plot + (richness_comp_plot / free(disease_conts) / disease_ord) +
  plot_layout(guides = "auto") +
  plot_annotation(tag_levels = "a") &
  theme(text = element_text(size = 6),
        title = element_text(size = 7))

ggsave("../Papers/Figures/disease_comparisons.png", dpi = 300, width = 180, height = 180, units = "mm")
```

## supplemental figures

host read proportions

```{r}
ps_organelles <- readRDS("data/16s_microbiome_phyloseq.rds") |> 
  psmelt() |> 
  filter(SampleType == "Sample") |> 
  mutate(
    Organelle = case_when(Family == "Mitochondria" ~ "Mitochondria",
                          Order == "Chloroplast" ~ "Plastid",
                          .default = "Bacteria")
  ) |> 
  mutate(Total = sum(Abundance), .by = Sample) |> 
  summarise(Abundance = sum(Abundance), 
            Total = unique(Total),
            Tissue = unique(Tissue),
            Site = unique(Site),
            Tree = unique(Tree),
            .by = c(Sample, Organelle)) |> 
  mutate(Prop = Abundance / Total)

organelles_plot <- ps_organelles |> 
  filter(Organelle != "Bacteria") |> 
  ggplot(aes(x = Tissue, 
             y = Prop)) +
  geom_point(aes(colour = Tissue),
             position = position_jitter()) +
  scale_colour_discrete(type = c("olivedrab3", "salmon4", "bisque2")) +
  scale_y_continuous(limits = c(0, 1)) +
  facet_grid(cols = vars(Organelle)) +
  labs(x = NULL, 
       y = "Proportion of reads") +
  theme_light() + 
  theme(legend.position = "none")

mg_ids <- tar_read(tree_df) |> 
  select(Leaf = MG_leaf, Stem = MG_stem, Soil = MG_soil) |> 
  pivot_longer(cols = everything(), names_to = "Tissue", values_to = "SampleID")

mg_input <- read_tsv(list.files("../Pipeline_outputs/Metagenome_QC/", pattern = "multiqc_fastqc_fastqc_raw.txt", recursive = TRUE, full.names = TRUE), id = "file") |> 
  mutate(SampleID = str_extract(Sample, "[:alnum:]+")) |> 
  select(SampleID, Count = `Total Sequences`) |> 
  summarise(Initial = sum(Count), .by = SampleID)

mg_fastp <- read_tsv(list.files("../Pipeline_outputs/Metagenome_QC/", pattern = "mqc_fastp_filtered_reads_plot_1.txt", recursive = TRUE, full.names = TRUE), id = "file") |> 
  mutate(SampleID = str_extract(Sample, "[:alnum:]+")) |> 
  select(SampleID, Count = `Passed Filter`) |> 
  summarise(FastP = sum(Count) / 2, .by = SampleID)

mg_hostfilter <- read_tsv(list.files("../Pipeline_outputs/Metagenome_QC/", pattern = "multiqc_fastqc_fastqc_trimmed.txt", recursive = TRUE, full.names = TRUE), id = "file") |> 
  mutate(SampleID = str_extract(Sample, "[:alnum:]+")) |> 
  select(SampleID, Count = `Total Sequences`) |> 
  summarise(Nonhost = sum(Count), .by = SampleID)

mg <- left_join(mg_input, mg_fastp) |> 
  left_join(mg_hostfilter) |> 
  left_join(mg_ids)

mg_stepplot <- mg |> 
  pivot_longer(cols = !c(SampleID, Tissue), names_to = "Step", values_to = "Reads") |> 
  ggplot(aes(x = fct_relevel(Step, c("Initial", "FastP", "Nonhost")))) +
  stat_lineribbon(aes(y = Reads, fill = after_stat(.width)), .width = ppoints(10)) +
  facet_grid(rows = vars(Tissue)) +
  scale_fill_distiller() +
  labs(x = "Processing stage",
       y = "Number of reads",
       fill = "Quantile") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

mg_props <- mg |> 
  mutate(Prop = Nonhost / FastP) |> 
  ggplot(aes(x = Tissue, y = 1 - Prop, colour = Tissue)) +
  geom_point(position = position_jitter()) +
  scale_colour_discrete(type = c("olivedrab3", "salmon4", "bisque2")) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = NULL, 
       y = "Proportion of reads") +
  theme_light() +
  theme(legend.position = "none")

prop_plot <- organelles_plot + mg_props + 
  plot_annotation(tag_levels = "a") + 
  plot_layout(axes = "collect")
```

Sequencing depth vs size factor value

```{r}
sf_vs_total <- function(marker, tissue) {
  ps <- tar_read_raw(glue::glue("microbiome_tj_corrected_{marker}_{tissue}"))
  sf <- tar_read_raw(glue::glue("model_draws_{marker}_{tissue}")) |> 
    subset_draws("SF") |> as_draws_rvars() |> pluck("SF")
  
  ps <- psmelt(ps) |> 
    arrange(Site, Tree) |> 
    summarise(Total = sum(Abundance), .by = c(Site, Tree)) |> 
    mutate(SF = sf,
           Marker = marker,
           Tissue = tissue)
}

sf_vs_total_df <- map(c("ITS", "16S"), \(x)
                      map(c("Leaf", "Stem", "Soil"), \(y)
                          sf_vs_total(x, y))
                      ) |> list_flatten() |> list_rbind()

sf_vs_total_plot <- ggplot(sf_vs_total_df, aes(x = Total, ydist = SF)) +
  stat_pointinterval(aes(colour = Tissue)) +
  scale_x_log10() +
  scale_colour_discrete(type = c("olivedrab3", "salmon4", "bisque2")) +
  facet_grid(rows = vars(Marker)) +
  labs(x = "Total reads sequenced in sample", y = "Model size factor") +
  theme_light()
```

dada2 plot

```{r}
its <- read_tsv("data/its_overall_summary.tsv")
rrna <- read_tsv("data/16s_overall_summary.tsv")

ps_its_summ <- tar_read(phyloseq_its) |> 
  psmelt() |> summarise(tax_filter = sum(Abundance), .by = c(Sample, SampleType)) |> 
  rename(sample = Sample)
ps_16s_summ <- tar_read(phyloseq_16s) |> 
  psmelt() |> summarise(tax_filter = sum(Abundance), .by = c(Sample, SampleType)) |> 
  rename(sample = Sample)

ps_data_summ <- bind_rows(ps_its_summ, ps_16s_summ)

dada2_data <- bind_rows(its, rrna) |> 
  mutate(Marker = str_extract(sample, "16S|ITS")) |> 
  select(-cutadapt_passing_filters_percent) |> 
  left_join(ps_data_summ)

dada2_plot <- dada2_data |> 
  pivot_longer(cols = !c(Marker, sample, SampleType), names_to = "Stage", values_to = "Reads") |> 
  filter(!is.na(Reads), SampleType == "Sample") |> 
  summarise(Reads = rvar(Reads), .by = c(Marker, Stage)) |> 
  mutate(median = median(Reads), stage_numeric = rep(1:9, times = 2)) |> 
  ggplot(aes(x = fct_reorder(Stage, stage_numeric))) +
  stat_lineribbon(aes(ydist = Reads, fill = after_stat(.width), group = Marker), .width = ppoints(10)) +
  facet_grid(rows = vars(Marker), scales = "free_y") +
  scale_fill_distiller() +
  #scale_y_continuous(transform = "log10") +
  labs(x = "Processing stage",
       y = "Number of reads",
       fill = "Quantile") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

metagenome assembly statistics

```{r}
assembly_files <- list.files("../Functional_analysis/data/quast", full.names = TRUE)
assembly_statistics <- map(assembly_files, 
                           \(x) read_tsv(x, id = "file") |> 
                             rename(Value = 3)) |> 
  list_rbind() |> 
  mutate(AssemblyID = str_extract(file, "\\.\\./Functional_analysis/data/quast/(.*)\\.tsv", group = 1),
         Tissue = substr(AssemblyID, 1, 4),
         Site = str_extract(AssemblyID, "([:ALPHA:]+)_", group = 1)) |> 
  select(Site, Tissue, Assembly, Value) |> 
  filter(Assembly %in% c("# contigs", 
                         "Largest contig",
                         "Total length",
                         "N50",
                         "L50",
                         "GC (%)",
                         "# predicted rRNA genes")) |> 
  rowwise() |> 
  mutate(Sample = glue::glue("{Tissue}_{Site}"),
                             temp = str_replace(Value, " part", ""), 
                             Value = eval(parse(text = temp))
         )

assembly_stats_plot <- ggplot(assembly_statistics, aes(x = Tissue, 
                                y = Value, 
                                colour = Tissue)) +
  geom_point(position = position_jitter(width = 0.3)) +
  facet_grid(rows = vars(Assembly), scales = "free") +
  labs(y = NULL)+
  theme_light() +
  theme(legend.position = "none")
```
