---
title: "Model predictions"
format:
    html:
        embed-resources: true
editor: visual
project:
  execute_dir: project
params:
  marker: "ITS"
  tissue: "Leaf"
---

```{r}
#| echo: false
#| include: false
library(targets)
library(tidyverse)
library(sf)
library(microViz)
library(patchwork)
library(phyloseq)
library(posterior)
library(cmdstanr)
library(ggdist)
library(patchwork)
source("R/model_plots.R")
```

Load the model and data:

```{r}
model_draws <- tar_read_raw(glue::glue("model_draws_{params$marker}_{params$tissue}")) |> 
  as_draws_rvars()
model_postpreds <- tar_read_raw(glue::glue("model_postpreds_{params$marker}_{params$tissue}")) |> 
  as_draws_rvars()
asv_table <- tar_read_raw(glue::glue("asv_table_{params$marker}_{params$tissue}"))
phylo <- tar_read_raw(glue::glue("microbiome_tj_corrected_{params$marker}_{params$tissue}"))

site <- tar_read(site_df) |> 
  mutate(Site_ID = 1:n())

tree <- tar_read(tree_df)|> 
  arrange(Site, Tree) |> 
  mutate(Tree_ID = 1:n()) |> 
  mutate(Disease = case_when(
                   str_detect(Tree, "C") ~ 0,
                   str_detect(Tree, "A") ~ 1,
                   str_detect(Tree, "E") ~ 2,
                   .default = NA
                 )
         )
```

## Alpha diversity

### Site-level

```{r}
#| echo: false
phylo_richness <- phylo |> 
  ps_select(Site, Tree) |> 
  psmelt() |> 
  left_join(site) |> 
  arrange(OTU, Site) |> 
  mutate(Ntrees = ifelse(AODProx == "AOD", 20, 5),
         SamplingDepth = sum(Abundance) / Ntrees,
         .by = c(Site)) |> 
  summarise(Richness = ifelse(sum(Abundance) > 0, 1, 0), 
            .by = c(Site, SamplingDepth, OTU)) |> 
  summarise(Richness_raw = sum(Richness), .by = c(Site, SamplingDepth)) |> 
  mutate(Site_num = as.character(1:n()))
  
site_richness <- model_postpreds |> 
  pluck("Z_cond") |> 
  rvar_apply(1, rvar_sum) |> 
  as_tibble() |> 
  rename(Richness_pred = value) |> 
  rownames_to_column("Site_num") |> 
  left_join(phylo_richness) |> 
  mutate(Site_num = as.numeric(Site_num),
         Richness_diff = Richness_pred - Richness_raw) |> 
  left_join(site)

ggplot(site_richness) +
  stat_pointinterval(aes(xdist = Richness_pred, 
                         y = fct_reorder(as.factor(Site), Northing),
                         colour = AODProx)) +
  geom_point(aes(x = Richness_raw, 
                 y = fct_reorder(as.factor(Site), Northing), 
                 colour = AODProx),
             shape = 4, size = 3) +
  theme_bw() +
  labs(x = "Species richness", y = "Site") + 
  theme(legend.position = "none") + 
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 18))

ggplot(site_richness) +
  stat_pointinterval(aes(ydist = Richness_diff, 
                         x = AODProx,
                         colour = AODProx),
                     position = position_dodge(width = 0.8)) +
  # facet_grid(rows = vars(AODProx), scales = "free_y") +
  theme_bw() +
  labs(y = "Difference between measured and predicted richness", x = "Mean reads per sample at site") + 
  theme(legend.position = "none")
```

### Sample-level

```{r}
#| echo: false
phylo_sample_richness <- phylo |> 
  ps_select(Site, Tree) |> 
  psmelt() |> 
  arrange(Site, Tree) |> 
  summarise(Richness_data = sum(Abundance > 0), 
            TotalReads = sum(Abundance),
            .by = c(Sample, Site, Tree)) |> 
  left_join(tree) |> 
  left_join(site) |> 
  mutate(Sample_num = 1:n())

sample_richness <- model_postpreds |> 
  pluck("U_cond") |> 
  rvar_apply(1, rvar_sum) |> 
  as_tibble() |> 
  rename(Richness_pred = value) |> 
  rownames_to_column("Sample_num") |> 
  mutate(Sample_num = as.numeric(Sample_num)) |> 
  left_join(phylo_sample_richness, by = "Sample_num") |> 
  mutate(Richness_diff = Richness_pred - Richness_data)

ggplot(sample_richness) +
  stat_pointinterval(aes(ydist = Richness_diff, 
                      x = TotalReads,
                      colour = as.factor(AODProx))) +
  scale_colour_discrete() +
  theme_bw() +
  labs(x = "Species richness", y = "Site", title = glue::glue("{params$tissue} {params$marker} ASVs"))
```

## Beta diversity

### Ordination - DCA

First: a visual comparison of community composition using ordination. To properly account for uncertainty in the outcome, we need to run an ordination over a whole drawset at once, and then do the posterior calculations afterwards, to ensure the scale is the same. Because of this, we will use DECORANA, as computing the distance matrix of a community matrix size ($40 \times 4000$) is computationally horrific.

To simplify this process somewhat, there are currently only 40 possible site-tree combinations in the model, so we will subset the data to each possible Site-Disease combination.

```{r}
#| echo: false

## subset data
non_duplicates_df <- sample_data(phylo) |> 
  as("data.frame") |> 
  mutate(RowID = 1:n()) |> 
  slice(1, .by = c(Site, Disease)) |> 
  left_join(site)

## get draws
Y_norm <- model_postpreds |> 
  pluck("Y_norm")

## subset draws using indices of subset data
Y_norm_nd <- Y_norm[non_duplicates_df$RowID,]

## reshape array to matrix for ordination
merged_Y_norm <- draws_of(Y_norm_nd) |> 
  apply(1, \(x) x, simplify = FALSE) |> 
  do.call("rbind", args = _)

## run ordination and get scores
bigdca <- vegan::decorana(merged_Y_norm)
bigscores <- vegan::scores(bigdca)

## reshape ordination scores back to array, convert to rvars
bigscores_rvar <- bigscores |> 
  array(dim = c(40, 4000, 2)) |> 
  aperm(c(2, 1, 3)) |> 
  rvar()

## plot modelled ordination
(model_ordination <- bigscores_rvar |> 
  as.data.frame() |> 
  bind_cols(non_duplicates_df) |> 
  rowwise() |> 
  mutate(mean_V1 = median(V1),
         mean_V2 = median(V2),
         lci_V1 = quantile(V1, probs = 0.975),
         lci_V2 = quantile(V2, probs = 0.975),
         uci_V1 = quantile(V1, probs = 0.025),
         uci_V2 = quantile(V2, probs = 0.025)
         ) |> 
  ggplot(aes(x = mean_V1, 
             y = mean_V2,
             xmin = lci_V1, xmax = uci_V1,
             ymin = lci_V2, ymax = uci_V2)
         ) +
  geom_errorbar(colour = "grey70") + 
  geom_errorbarh(colour = "grey70") +
  geom_point(aes(colour = AODProx), size = 3) +
  # scale_color_viridis_c() +
  labs(x = "DCA1", y = "DCA2") +
  theme_bw())
```

## Occupancy predictor summaries

### Environmental sigmas

```{r}
model_draws |> subset_draws("sigmas_occ") |> 
  bayesplot::mcmc_dens()
```

### Disease sigmas

```{r}
model_draws |> subset_draws("sigmas_rep") |> 
  bayesplot::mcmc_dens()
```

### Per-species AOD response

```{r}
intervals <- disease_intervals(model_draws, disease_col = 12)
plot_disease_intervals(intervals)
```

```{r}
filter(intervals, uci < 1) |> 
  left_join(asv_table) |> 
  arrange(median) |> 
  select(Phylum, Genus, Species, median) |> 
  mutate(median = round(median, digits = 2)) |> 
  head(n = 10) |> 
  gt::gt() |> 
  gt::cols_label(
    median = "Change in odds ratio"
  )
```

## Occupancy slopes

```{r}
groundfrost <- plot_species_slopes(model_draws, site$groundfrost, 2, 2) + 
  labs(x = "Mean annual days of groundfrost",
       title = "Groundfrost")
hurs <- plot_species_slopes(model_draws, site$hurs, 3, 3) + 
  labs(x = "Mean annual relative humidity (%)",
       title = "Humidity")
pv <- plot_species_slopes(model_draws, site$pv, 4, 4) + 
  labs(x = "Mean annual vapour pressure (hPa)",
       title = "Vapour pressure")
rainfall <- plot_species_slopes(model_draws, site$rainfall, 5, 5) + 
  labs(x = "Mean annual rainfall (mm)",
       title = "Rainfall")
sfcWind <- plot_species_slopes(model_draws, site$sfcWind, 6, 6) + 
  labs(x = "Mean annual wind speed (knots)",
       title = "Windspeed")
sun <- plot_species_slopes(model_draws, site$sun, 7, 7) + 
  labs(x = "Duration of bright sunshine (hours)",
       title = "Sunshine")
tas <- plot_species_slopes(model_draws, site$tas, 8, 8) + 
  labs(x = "Mean annual temperature (C)",
       title = "Temperature")
Ntot <- plot_species_slopes(model_draws, site$Ntot, 9, 9) + 
  labs(x = "Mean annual N deposition (kg/ha/y)",
       title = "Nitrogen deposition")
SOx <- plot_species_slopes(model_draws, site$SOx, 10, 10) + 
  labs(x = "Mean annual SOx deposition (kg/ha/y)",
       title = "Sulphur deposition")
CaMg <- plot_species_slopes(model_draws, site$CaMg, 11, 11) + 
  labs(x = "Mean annual Ca/Mg deposition (kg/ha/y)",
       title = "Ca/Mg deposition")

groundfrost + hurs + pv + rainfall + sfcWind +sun + tas + Ntot+ SOx + CaMg + plot_layout(nrow = 2, axes = "collect_y") & labs(x = "") & coord_cartesian(y = c(0, 12))
```
