---
project:
  execute-dir: project
format:
    html:
        embed-resources: true
        default-image-extension: svg
params:
  tissue: "Leaf"
  marker: "ITS"
---

# Model diagnostics for `r params$tissue` `r params$marker` 3-level occupancy model.

## Load the model and packages

```{r}
#| echo: false
library(posterior)
library(tidyverse)
library(bayesplot)
library(cmdstanr)
library(targets)

phylo <- tar_read_raw(glue::glue("microbiome_filtered_{params$marker}_{params$tissue}"))
model_summary <- tar_read_raw(glue::glue("model_summary_{params$marker}_{params$tissue}"))
model_diagnostics <- tar_read_raw(glue::glue("model_diagnostics_{params$marker}_{params$tissue}"))
model_draws <- tar_read_raw(glue::glue("model_draws_{params$marker}_{params$tissue}")) |> 
  as_draws_rvars()
model_postpreds <- tar_read_raw(glue::glue("model_postpreds_{params$marker}_{params$tissue}")) |> 
  as_draws_rvars()
model_data <- tar_read_raw(glue::glue("stan_data_{params$marker}_{params$tissue}"))
loo <- tar_read_raw(glue::glue("loo_diag_{params$marker}_{params$tissue}"))
```

## MCMC diagnostics

Basic MCMC diagnostic checks. Models with divergent transitions, or large numbers of max_treedepth errors, are likely problematic and need refinement.

```{r}
#| echo: false
model_diagnostics
```

### Rhats

Rhat is a parameter-level summary statistic that attempts to summarise the convergence of a set of HMC/MCMC chains. A value close to 1 (approximately < 1.05) indicates chain convergence for that parameter. Below is a histogram of Rhat values for all model parameters, giving a rough at-a-glance summary of model convergence.

```{r}
#| echo: false
model_summary |> 
  ggplot(aes(x = rhat)) +
  geom_histogram() +
  labs(x = "Rhat", y = "Frequency") +
  theme_bw()
```

It should be noted that while this is a good rough guide, it is still important to inspect the MCMC chains more closely to gain further assurances of convergence.

### Trace plots

Below are MCMC traceplots for a selection of parameters of different types, to give a better overview of any convergence problems that may have occurred.

#### Sigmas

Evironment sigmas:

```{r}
#| echo: false
posterior::subset_draws(model_draws, "sigmas_env") |> bayesplot::mcmc_trace()
```

Sample sigmas:

```{r}
#| echo: false
posterior::subset_draws(model_draws, "sigmas_sample") |> bayesplot::mcmc_trace()
```

#### Proportions

Environment props:

```{r}
#| echo: false
pluck(model_draws, "props_env") |> {\(x) x[,1]}() |> as_draws_array() |> bayesplot::mcmc_trace()
pluck(model_draws, "props_env") |> {\(x) x[,2]}() |> as_draws_array() |> bayesplot::mcmc_trace()
pluck(model_draws, "props_env") |> {\(x) x[,3]}() |> as_draws_array() |> bayesplot::mcmc_trace()
```

Sample props:

```{r}
#| echo: false
pluck(model_draws, "props_sample") |> {\(x) x[,1]}() |> as_draws_array() |> bayesplot::mcmc_trace()
pluck(model_draws, "props_sample") |> {\(x) x[,2]}() |> as_draws_array() |> bayesplot::mcmc_trace()
```

#### Fixed effects

Site occupancy:

```{r}
#| echo: false
pluck(model_draws, "b_env") |> {\(x) x[1,]}() |> as_draws_array() |> bayesplot::mcmc_trace()
pluck(model_draws, "b_env") |> {\(x) x[2,]}() |> as_draws_array() |> bayesplot::mcmc_trace()
pluck(model_draws, "b_env") |> {\(x) x[3,]}() |> as_draws_array() |> bayesplot::mcmc_trace()
```

Sample occupancy:

```{r}
#| echo: false
pluck(model_draws, "b_sample") |> {\(x) x[1,]}() |> as_draws_array() |> bayesplot::mcmc_trace()
pluck(model_draws, "b_sample") |> {\(x) x[2,]}() |> as_draws_array() |> bayesplot::mcmc_trace()
```

#### Species effects

Choose three random species to evaluate for:

```{r}
n_spp <- pluck(model_draws, "b_env_species") |> {\(x) dim(x)[3]}()
spp <- sample(n_spp, size = 3)

spp
```

Environment effects:

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 16
pluck(model_draws, "b_env_species") |>
  {\(x) drop(x[1,,])}() |> 
  draws_rvars(b_env_species = _) |> 
  bayesplot::mcmc_trace(pars = vars(param_glue("b_env_species[{var},{species}]",
                                               var = c(1:11),
                                               species = spp)
                                    )
                        )

pluck(model_draws, "b_env_species") |>
  {\(x) drop(x[2,,])}() |> 
  draws_rvars(b_env_species = _) |> 
  bayesplot::mcmc_trace(pars = vars(param_glue("b_env_species[{var},{species}]",
                                               var = c(1:11),
                                               species = spp)
                                    )
                        )

pluck(model_draws, "b_env_species") |>
  {\(x) drop(x[3,,])}() |> 
  draws_rvars(b_env_species = _) |> 
  bayesplot::mcmc_trace(pars = vars(param_glue("b_env_species[{var},{species}]",
                                               var = c(1:11),
                                               species = spp)
                                    )
                        )
```

Sample occupancy:

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 16
pluck(model_draws, "b_sample_species") |>
  {\(x) drop(x[1,,])}() |> 
  draws_rvars(b_sample_species = _) |> 
  bayesplot::mcmc_trace(pars = vars(param_glue("b_sample_species[{species}]",
                                               species = spp)
                                    )
                        )

pluck(model_draws, "b_sample_species") |>
  {\(x) drop(x[2,,])}() |> 
  draws_rvars(b_sample_species = _) |> 
  bayesplot::mcmc_trace(pars = vars(param_glue("b_sample_species[{species}]",
                                               species = spp)
                                    )
                        )
```

#### Dispersion

```{r}
#| echo: false
posterior::subset_draws(model_draws, c("sigma_phi", "a_phi")) |> bayesplot::mcmc_trace()
posterior::subset_draws(model_draws, c("z_phi")) |> bayesplot::mcmc_trace(pars = vars(param_range("z_phi", spp)))
```

#### Size factors

```{r}
n_size_factors <-  posterior::subset_draws(model_draws, "SF") |> as_draws_rvars() |> pluck("SF") |> length()
sfs <- sample(n_size_factors, 10)

sfs

posterior::subset_draws(model_draws, c("SF")) |> bayesplot::mcmc_trace(pars = vars(param_range("SF", sfs)))
```

## Leave-one-out cross-validation

Assuming MCMC diagnostics have passed, below is the leave-one-out cross validation score for the model, with moment matching.

```{r}
n_pars <- model_summary |> 
  filter(!str_detect(variable, "species")) |> 
  filter(variable != "phi") |> 
  filter(variable != "log_lik")

n_pars

loo$estimates
```

## Goodness-of-fit

```{r}
Y <- model_data$Y |> as.vector()
Y_rep <- model_postpreds$Y_rep |> 
  t() |> 
  as.vector()

len_Y <- length(Y)
max_sample <- 5000
if(len_Y > max_sample) {
  which_Y <- sample(1:len_Y, size = max_sample)
} else {
  which_Y <- 1:len_Y
}

bayesplot::ppc_ecdf_overlay(y = Y[which_Y],
                    yrep = draws_of(Y_rep[which_Y]), 
                    discrete = TRUE
                    ) +
  scale_x_continuous(trans = "pseudo_log")

bayesplot::ppc_rootogram(y = Y[which_Y],
                    yrep = draws_of(Y_rep[which_Y]), 
                    style = "hanging"
                    ) +
  scale_x_continuous(trans = "pseudo_log")
```
